---
title: "Latent non-Gaussian models with ngvb2"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    highlight: tango
    theme:
      version: 5
      bootswatch: flatly
  rmarkdown::html_vignette: default
vignette: "%\\VignetteIndexEntry{Latent non-Gaussian models with ngvb2} %\\VignetteEngine{knitr::rmarkdown}
  %\\VignetteEncoding{UTF-8}\n"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 8, fig.height = 3.6, fig.align = "center",
                      dpi = 96, message = FALSE, warning = FALSE,
                      eval = requireNamespace("INLA", quietly = TRUE))
library(ggplot2)
have_areal   <- all(vapply(c("spdep", "sf", "spData"), requireNamespace,
                           logical(1), quietly = TRUE)) &&
                requireNamespace("INLA", quietly = TRUE)
have_fmesher <- requireNamespace("fmesher", quietly = TRUE) &&
                requireNamespace("INLA", quietly = TRUE)
set.seed(1)
```

```{r lib}
# devtools::install("C:/Users/rafae/Documents/ngvb/ngvb2")  # local dev only; do not run at build time
library(ngvb2)
```

# Overview

A latent Gaussian model (LGM), the workhorse of R-INLA, places a Gaussian prior on a
latent field: a random walk, an autoregression, a spatial effect, or a set of random
effects. That prior is convenient but rigid, because it assumes the same variability
everywhere. Real data often break the assumption in just a few places, such as a sudden
jump in a time series, one outlying subject, a sharp boundary between regions, or a local
hotspot in space. A Gaussian model then has to compromise. It either over-smooths the
interesting feature or inflates the variance everywhere to accommodate it.

`ngvb2` relaxes that single assumption. It keeps everything you already have, the same
INLA model and the same code, and replaces the Gaussian driving noise with a heavier-tailed
one. The result adds flexibility only where the data ask for it and falls back to the
Gaussian model everywhere else. Two functions do the work, and both take an ordinary `inla`
fit that was run with `control.compute = list(config = TRUE)`:

* `ng.check(fit)` asks whether you should relax the Gaussian assumption, and where.
* `ngvb(fit)` fits the non-Gaussian extension.

Both auto-detect the models in your fit. Out of the box that covers `iid` random effects,
random walks `rw1` and `rw2`, autoregressions `ar1`, intrinsic areal models `besag`, Matern
(`SPDE`) fields, and `generic0` structure matrices. Anything else can be supplied through
`ngvb_custom()`, described in the last section.

Every example below follows the same short recipe: plot the data, write down the model,
fit the Gaussian version with INLA, check it with `ng.check()`, and then extend it with
`ngvb()`.

# The model in one paragraph

Every LGM writes its latent field $\mathbf x$ through a model-specific dependency matrix
$\mathbf D(\boldsymbol\theta)$ that turns $\mathbf x$ into independent Gaussian increments,
$$\mathbf D(\boldsymbol\theta)\,\mathbf x \;\overset{d}{=}\; \boldsymbol\Lambda,
\qquad \Lambda_i \sim N(0,\,h_i).$$
For a random walk $\Lambda_i=x_{i}-x_{i-1}$, for an autoregression
$\Lambda_i=x_i-\rho x_{i-1}$, and so on. The non-Gaussian model keeps the same
$\mathbf D$ but lets each increment have its own variance $V_i$,
$$\Lambda_i \mid V_i \sim N(0,\,V_i),\qquad
  V_i \sim \mathrm{IG}(h_i,\,\eta^{-1}h_i^2),\qquad E[V_i]=h_i .$$
The mixing variables $\mathbf V$ carry the local flexibility. Where $V_i=h_i$ the model is
Gaussian, and where $V_i$ is inflated the increment is allowed to be unusually large. A
single non-Gaussianity parameter $\eta\ge 0$ controls how far $\mathbf V$ may stray from
$\mathbf h$, and $\eta=0$ is exactly the Gaussian model. Conditional on $\mathbf V$ the
field is still Gaussian, with precision
$$\mathbf Q(\boldsymbol\theta,\mathbf V)=\mathbf D(\boldsymbol\theta)^\top\,
   \mathrm{diag}(1/\mathbf V)\,\mathbf D(\boldsymbol\theta).$$
`ngvb2` fits the model with a short variational-Bayes loop that repeatedly calls INLA with
$\mathbf V$ fixed and then updates $\mathbf V$ and $\eta$. The prior on $\eta$ is an
exponential (a penalised-complexity prior with rate `alpha.eta`, default $2$) that shrinks
toward the Gaussian model, so non-Gaussianity has to earn its place. Increase `alpha.eta`
for stronger shrinkage.

Two variational algorithms are available. The default, `method = "SVI"`, is a structured
mean-field scheme that updates $q(\mathbf V)$ and $q(\eta)$ in turn, and it is the more
reliable choice. The alternative, `method = "SCVI"`, integrates $\mathbf V$ out of the
$\eta$ update analytically and so converges in fewer iterations, but its collapsed marginal
can over-shrink a weak yet genuine effect all the way to the Gaussian model. The two agree
when the signal is strong or clearly absent, and they disagree in the weak-signal regime, so
prefer SVI unless you have checked that SCVI gives the same answer on your problem.

There is one more thing worth knowing before the examples. The conditional precision
$\mathbf Q=\tau\,\mathbf D_0^\top\mathrm{diag}(1/\mathbf V)\mathbf D_0$ is exactly invariant
under $(\tau,\mathbf V)\mapsto(c\tau,c\mathbf V)$, meaning the overall scale can be carried
either by the precision $\tau$ or by the mixing variables. Nothing in the data distinguishes
the two, only the prior does, so on weak or homogeneous signal the loop can drift along that
flat ridge, let $\mathbf V$ absorb the scale, and inflate $\eta$ spuriously. `ngvb()` closes
the ridge by default with `identify.scale = TRUE`, which re-anchors each component's
discrepancies every iteration so their total matches the Gaussian total. That pins the scale
to $\tau$ and leaves $\eta$ to respond only to the relative pattern of the increments, which
is the genuine-outlier information it should measure. Well-scaled fits are left essentially
untouched.

# Which models and likelihoods are supported

Running `inla.list.models()` prints everything R-INLA offers: dozens of latent model
components under `[latent]`, and a long list of `[likelihood]` families. The non-Gaussianity
in `ngvb2` lives entirely in the latent field, on its driving noise, and never touches the
observation model. So **every INLA likelihood is available**: Gaussian, Poisson, binomial,
negative binomial, survival families, and the rest all work unchanged. You just fit your LGM
as usual and hand it to `ng.check()` or `ngvb()`.

The latent components split into three groups.

**Detected automatically.** If your fit uses one of these, `ngvb()` and `ng.check()` find it
with no extra arguments: `iid`, `rw1`, `rw2`, `ar1`, `besag` (the intrinsic CAR), `spde`
/ `spde2` (Matern fields), and `generic0` (any user-supplied structure matrix `Cmatrix`).
Each of these is a single operator applied to independent noise, so it has the dependency
matrix $\mathbf D$ the method needs.

**Added by hand.** Some models are not auto-detected but are still of the operator form, so
you build them with `ngvb_operator()` or `ngvb_custom()` and pass them through `components`.
This covers the simultaneous autoregression (`sar`), a proper row-standardised CAR (`car`),
the Ornstein-Uhlenbeck process (`ou`), a `seasonal` operator, and AR($p$). The `generic0`
route is the general escape hatch: any Gaussian component whose precision is a fixed structure
matrix times a scale (for instance `crw2`, `rw2d`, `matern2d`, `dmatern`) can be fit by
passing that matrix as `Cmatrix`.

**Not compatible.** Some components are not a single operator applied to one field, and the
method does not apply to them. The main cases are the additive spatial models `bym`, `bym2`,
`besag2` (a latent field that is a sum of parts), the conditionally specified `besagproper`,
the copy and mixture constructions `generic1`, `generic2`, `generic3`, `copy`, `scopy`, `z`,
the measurement-error models `mec`, `meb`, and the nonlinear-covariate effects `sigm`,
`revsigm`, `log1exp`, `logdist`. For an additive model such as `bym`, the practical fix is to
split it into its `besag` and `iid` parts as two separate `f()` terms, which `ngvb()` then
extends jointly.

# A time series with jumps

`jumpts` is a series that is smooth apart from two abrupt jumps.

```{r rw1-data, fig.height = 3}
ggplot(jumpts, aes(x, y)) +
  geom_line(colour = "grey70") +
  geom_point(colour = "grey30", size = 1.3) +
  labs(title = "jumpts", x = "time", y = "y") +
  theme_minimal(base_size = 12)
```

A first-order random walk models the series through its increments,
$$x_i - x_{i-1} \sim N(0,\,1/\tau),$$
which is the operator $\mathbf D = \sqrt{\tau}\,\mathbf D_0$ with $\mathbf D_0$ the
first-difference matrix. A Gaussian RW1 forces every increment to share the same variance,
so it cannot tell a jump from ordinary noise. Fit it with INLA:

```{r rw1-inla}
LGM <- inla(y ~ -1 + f(x, model = "rw1"), data = jumpts,
            control.compute = list(config = TRUE))
```

Now check whether the latent Gaussian assumption holds:

```{r rw1-check, fig.height = 3.4}
ng.check(LGM, compute.fixed = FALSE)
```

The left panel is the per-index sensitivity $d_i(y)$, which measures how much the evidence
would improve if increment $i$ were allowed to be non-Gaussian. It spikes at the two jumps.
The right panel compares the observed overall sensitivity $s_0=\sum_i d_i$ (the vertical
line) against the distribution it would have under a truly Gaussian field (the grey curve).
The observed value sits far out in the tail, with $p\approx0$, so the Gaussian RW1 is
inadequate and we know exactly where.

Extend it. The model is auto-detected, so this is a single call:

```{r rw1-fit, fig.height = 6}
LnGM <- ngvb(LGM)
plot(LnGM)
```

For each component the plot shows the mixing weights $V_i/h_i$ (the posterior mean with a
90 percent credible interval, and a cross at the plug-in value $1/E[1/V_i]$ that INLA is
actually refit with, which sits a little below the mean because $q(V_i)$ is right-skewed),
next to the posterior density of $\eta$ (mean solid, median dashed). The final panel tracks
$E[\eta]$ across the iterations. The mixing variables are inflated precisely at the two
jumps and sit near one elsewhere, so the model bought local flexibility exactly where it was
needed and stayed Gaussian in between. The non-Gaussianity parameter is clearly positive:

```{r rw1-eta}
LnGM$eta
```

How much better is the non-Gaussian model overall? We answer that with a Bayes factor.
`ngvb_sample()` draws the mixing vector $\mathbf V$ from its posterior, refits INLA at each
draw, and `bayes.factor()` compares the marginal likelihood of the non-Gaussian model against
the Gaussian one with $\mathbf V$ integrated out rather than fixed at a plug-in value. Because
$\mathbf V$ is averaged over, the comparison stays honest even at small sample sizes.

```{r rw1-bf}
bayes.factor(LnGM, n.samples = 30, seed = 1)
```

On the usual scale a $\log_{10}$ Bayes factor above $0.5$, $1$, or $2$ is substantial, strong,
or decisive evidence for the non-Gaussian model. Here it is far into the decisive range. The
`ess` field is the effective number of draws behind the estimate; if it is small, raise
`n.samples`.

# Random intercepts and slopes

`Orthodont` records a growth measurement for 27 children at four ages, together with a sex
indicator and time coded two ways. A quick look at the raw data frame:

```{r ortho-data, fig.height = 5}
plot(Orthodont)
```

Two kinds of child-to-child variation matter here. Each child starts at a different height,
which is variation in the intercept, and each child grows at a slightly different rate, which
is variation in the slope. We give every child its own intercept and its own slope, as two
`iid` random effects:
$$\text{value}_{ij} = \beta_0 + \beta\,\text{covariates}
   + \underbrace{a_{\text{subject}(ij)}}_{\text{random intercept}}
   + \underbrace{b_{\text{subject}(ij)}\,\text{time}_{ij}}_{\text{random slope}}
   + \varepsilon_{ij},$$
with $a_k \sim N(0, 1/\tau_a)$ and $b_k \sim N(0, 1/\tau_b)$ across children.

```{r ortho-inla}
formula <- value ~ 1 + Female + time + tF +
  f(subject,  model = "iid") +      # random intercept, one per child
  f(subject2, time, model = "iid")  # random slope, one per child

LGM <- inla(formula, data = Orthodont, control.compute = list(config = TRUE))
```

Why make the random effects non-Gaussian? A Gaussian random effect treats the children as
exchangeable draws from one normal distribution. If a child or two are genuine outliers,
with an unusually high baseline for example, the Gaussian either shrinks them toward the
mean and hides the outlier, or it widens the distribution for everyone to accommodate them.
A non-Gaussian random effect keeps a tight distribution for the typical children and lets
the few outliers stand out. `ng.check()` reports one diagnostic per component:

```{r ortho-check, fig.height = 6}
ng.check(LGM, compute.fixed = FALSE)
```

```{r ortho-ngvb, fig.height = 8.5}
LnGM <- ngvb(LGM)
plot(LnGM)
```

```{r ortho-eta}
LnGM$eta          # one non-Gaussianity parameter per component
```

The intercept carries more non-Gaussianity than the slope. A couple of children have
outlying baselines, while the growth rates are close to Gaussian. Nothing was specified by
hand: the prior on $\eta$ shrank the slope component back toward Gaussian on its own and
left the intercept free to flag its outliers.

```{r ortho-bf}
bayes.factor(LnGM, n.samples = 30, seed = 1)
```

# Areal data: an intrinsic model (besag)

Residential burglary rates in 49 counties of Columbus, Ohio, come with the `spData` package.
Plot the response on the map first:

```{r besag-data, eval = have_areal, fig.height = 4.2, fig.width = 5.5}
library(sf); library(spdep)
map  <- st_read(system.file("shapes/columbus.gpkg", package = "spData"), quiet = TRUE)
areal.plot(map, map$CRIME, title = "crime")
```

An intrinsic conditional autoregression (the `besag` model) assumes neighbouring counties
are similar, with Gaussian differences across the edges of the adjacency graph,
$$x_i - x_j \sim N(0,\,1/\tau)\quad\text{for each edge }(i,j).$$
The non-Gaussian version allows a few sharp boundaries between otherwise smooth regions.
Build the adjacency and fit the Gaussian model:

```{r besag-inla, eval = have_areal}
d    <- st_drop_geometry(map[, c("CRIME", "HOVAL", "INC")])
d$s  <- seq_len(nrow(d))
nb   <- poly2nb(map)
Wadj <- Matrix::sparseMatrix(i = rep(seq_along(nb), lengths(nb)), j = unlist(nb), x = 1)

LGM <- inla(CRIME ~ 1 + HOVAL + INC +
              f(s, model = "besag", graph = Wadj, scale.model = TRUE),
            data = d, control.compute = list(config = TRUE))
```

```{r besag-check, eval = have_areal, fig.height = 3.4}
ng.check(LGM, compute.fixed = FALSE)   # areal model auto-detected
```

The 118 edges of this graph all show a similarly small discrepancy from the Gaussian
reference. Exactly one exceeds twice the median and none exceeds three times it, whereas one
of the `jumpts` increments earlier was about 500 times its median. This field is essentially
Gaussian, and there is no sharp boundary to find. It is also exactly the homogeneous pattern
that could let $\eta$ and the precision run up together, so the scale-identifiability
constraint (on by default) matters here. It holds the fit to the honest answer near
$\eta\approx0.1$, with every mixing weight sitting at $\mathbf h$:

```{r besag-fit, eval = have_areal, fig.height = 6}
LnGM <- ngvb(LGM)                        # identify.scale = TRUE by default
plot(LnGM)
```

```{r besag-eta, eval = have_areal}
LnGM$eta                                # for besag, V lives on the graph edges
```

Turning the constraint off with `ngvb(LGM, identify.scale = FALSE)` would instead let $\eta$
drift up to around three while pushing every edge weight below $\mathbf h$. That uniform
shrinkage is the signature of a scale runaway rather than a real finding, and `ngvb()` would
warn about it.

For besag the mixing weights live on the graph edges, not the counties, so to see the spatial
pattern we average the incident-edge weights $V_i/h_i$ onto each county and map that. As
expected for this near-Gaussian field, the map is almost flat at one:

```{r besag-map, eval = have_areal, fig.height = 4.2, fig.width = 5.5}
inc      <- abs(LnGM$ops$s$Dfunc(LnGM$ops$s$theta.initial)) > 0   # edge-by-county incidence
countyVh <- as.numeric(Matrix::t(inc) %*% (LnGM$V$s / LnGM$h$s)) / Matrix::colSums(inc)
areal.plot(map, countyVh, title = "mean V/h")
```

```{r besag-bf, eval = have_areal}
bayes.factor(LnGM, n.samples = 30, seed = 1)
```

# Geostatistical data: a Matern (SPDE) field

`weatherdata`, bundled with the package, holds pressure readings across the Pacific
Northwest. Build the mesh first and plot the stations over it:

```{r spde-data, eval = have_fmesher, fig.height = 4.2, fig.width = 5.5}
w    <- read.csv(system.file("extdata", "weatherdata.csv", package = "ngvb2"))
mesh <- inla.mesh.2d(loc = w[, c("lon", "lat")], max.edge = c(1.5, 2.5),
                     cutoff = 0.3, max.n.strict = 400)
geo.plot(w[, c("lon", "lat", "press")], mesh = mesh, title = "pressure",
         palette = "RdBu")
```

We model pressure as a continuous Matern field through the SPDE approach, the standard tool
for point-referenced spatial data. The field solves a stochastic PDE whose finite-element
form gives the operator $\mathbf D=\kappa^2\mathbf C+\mathbf G$, where $\mathbf C$ and
$\mathbf G$ are the mass and stiffness matrices of the mesh. A Gaussian Matern smooths the
whole surface uniformly, while the non-Gaussian version lets a few locations depart more
sharply. `ngvb2` reads $\mathbf D$ straight from the fitted `spde` object, so you pass
nothing extra.

```{r spde-inla, eval = have_fmesher}
spde <- inla.spde2.matern(mesh, alpha = 2)
A    <- inla.spde.make.A(mesh, loc = as.matrix(w[, c("lon", "lat")]))
stk  <- inla.stack(tag = "est", data = list(y = w$press), A = list(A),
                   effects = list(s = 1:spde$n.spde))

LGM <- inla(y ~ -1 + f(s, model = spde), data = inla.stack.data(stk),
            control.predictor = list(A = inla.stack.A(stk)),
            control.compute = list(config = TRUE))
```

```{r spde-check, eval = have_fmesher, fig.height = 3.4}
ng.check(LGM, compute.fixed = FALSE)   # SPDE model recovered from the fitted object
```

```{r spde-fit, eval = have_fmesher, fig.height = 6}
LnGM <- ngvb(LGM, verbose = FALSE)
plot(LnGM)
```

```{r spde-eta, eval = have_fmesher}
LnGM$eta
```

The mixing variables live on the mesh nodes, so we can map them back over space. We plot the
ratio $V_i/h_i$, which is one where the field is Gaussian and larger where it departs from a
uniform Matern (for the SPDE the weights $h_i$ are the mesh mass, so $V_i/h_i$ rather than
$V_i$ is the quantity to look at):

```{r spde-map, eval = have_fmesher, fig.height = 4.2, fig.width = 5.5}
nodes <- data.frame(lon = mesh$loc[, 1], lat = mesh$loc[, 2],
                    Vh = LnGM$V$s / LnGM$h$s)
geo.plot(nodes, mesh = mesh, title = "V/h")
```

```{r spde-bf, eval = have_fmesher}
bayes.factor(LnGM, n.samples = 30, seed = 1)
```

`ngvb2` recognises the SPDE component because its `f(...)` model is an object built by
`inla.spde2.matern()`. It evaluates that argument from the fit's formula and reads the
finite-element matrices back from it.

# Your own precision matrix with ngvb_custom

If a model is not built in, you can supply the dependency matrix
$\mathbf D(\boldsymbol\theta)$ and the constant vector $\mathbf h$ yourself with
`ngvb_custom()`. The engine then forms
$\mathbf Q(\mathbf V)=\mathbf D^\top\mathrm{diag}(1/\mathbf V)\mathbf D$, owns its normalizing
constant, and `ng.check()` and `ngvb()` work exactly as before.

For areal data the natural custom model is the simultaneous autoregression (SAR), which uses
$$\mathbf D_{\mathrm{SAR}}=\sqrt{\tau}\,(\mathbf I-\rho\mathbf W),\qquad |\rho|<1,$$
with $\mathbf W$ a row-standardised adjacency. This is worth stressing: among areal models,
the SAR is the one you build through `ngvb_custom()`. A proper CAR, or the intrinsic
besag/ICAR family, cannot be written as a custom $\mathbf D$ of this kind, because their
precisions are specified conditionally or intrinsically rather than through a whitening
operator with a fixed sparsity pattern. The intrinsic besag model is instead available by
auto-detection, as in the previous section, and there is no way to hand it to `ngvb_custom()`
as a bespoke operator. So for a custom areal model, think SAR.

Unlike the intrinsic besag effect, the SAR mixing variables live on the counties themselves,
so we can map them directly.

Here is the full definition, and each argument is explained just below it.

```{r sar-build, eval = have_areal}
W      <- Matrix::Diagonal(x = 1 / rowSums(Wadj)) %*% Wadj    # row-standardise
N      <- nrow(W)
eigenv <- Re(eigen(as.matrix(W), only.values = TRUE)$values)

op_sar <- ngvb_custom(
  D = function(theta) {
    rho <- exp(theta[2]) / (1 + exp(theta[2]))
    sqrt(exp(theta[1])) * (Matrix::Diagonal(N) - rho * W)
  },
  h        = rep(1, N),
  ntheta   = 2,
  logprior = function(theta) {
    rho <- exp(theta[2]) / (1 + exp(theta[2]))
    stats::dnorm(theta[1], 0, 3, log = TRUE) + log(rho) + log(1 - rho)
  },
  lognc = function(theta, Vinv) {
    rho <- exp(theta[2]) / (1 + exp(theta[2]))
    0.5 * N * theta[1] + sum(log(1 - rho * eigenv)) +
      0.5 * sum(log(Vinv)) - 0.5 * N * log(2 * pi)
  })
```

The arguments are:

* `D = function(theta)` builds the dependency matrix $\mathbf D(\boldsymbol\theta)$ as a
  sparse `Matrix`. Its rows are the driving-noise terms and its columns are the latent nodes.
  Here `theta[1]` is the log precision and `theta[2]` is the internal (unconstrained) version
  of $\rho$, mapped into $(0,1)$ by the logistic function. INLA works on this internal scale,
  which is why the transform appears inside `D`.
* `h` is the constant vector with $E[V_i]=h_i$, and $V=h$ recovers the Gaussian model. For a
  discrete areal model there is one innovation per node, so `h` is a vector of ones.
* `ntheta` is the number of hyperparameters that INLA estimates, here two.
* `logprior = function(theta)` is the log prior density of the hyperparameters, written on
  the internal scale and including any change-of-variables Jacobian. This is the piece INLA
  adds to the log-posterior for $\boldsymbol\theta$. In this example there is a
  $N(0,3)$ prior on the log precision, and a uniform prior on $\rho\in(0,1)$; after the
  logistic transform the uniform density contributes the Jacobian term
  $\log\rho + \log(1-\rho)$. If you omit `logprior`, an independent $N(0,3)$ on each internal
  hyperparameter is used.
* `lognc = function(theta, Vinv)` is optional but recommended when $\mathbf D$ can get close
  to singular, as it does here when $\rho$ approaches one. It returns the log normalizing
  constant of the conditional Gaussian density,
  $\tfrac12\log|\mathbf Q(\boldsymbol\theta,\mathbf V)| - \tfrac{n}{2}\log(2\pi)$, and it
  receives `Vinv` $=1/\mathbf V$ so that the $\mathbf V$-dependent part of the determinant is
  included. For the SAR the log-determinant has the closed form
  $N\log\tau + 2\sum_i\log(1-\rho\lambda_i) + \sum_i\log(1/V_i)$, using the eigenvalues
  $\lambda_i$ of $\mathbf W$, which is both fast and stable near the boundary. When `lognc` is
  omitted, the engine computes the determinant generically, which is fine for well-conditioned
  operators.

A couple of practical notes. `D`, `logprior`, and `lognc` are evaluated in a separate R
process where only the base packages are attached, so any non-base function must be
namespace-qualified, which is why the code writes `Matrix::Diagonal` and `stats::dnorm`, and
uses `exp(t)/(1+exp(t))` rather than `plogis`. The data those functions reference, such as
`W` and `eigenv`, is captured automatically.

With the operator defined, fit the Gaussian SAR through the same engine and then extend it:

```{r sar-fit, eval = have_areal, fig.height = 6}
LGM  <- inla(CRIME ~ 1 + HOVAL + INC + f(s, model = ngvb_rgeneric(op_sar)),
             data = d, control.compute = list(config = TRUE))
LnGM <- ngvb(LGM, components = list(s = op_sar), verbose = FALSE)
plot(LnGM)
```

Because the SAR weights live on the counties, we can put them straight back on the map. Low
values are yellow and high values are red, so the reddest counties are where the model added
the most flexibility, meaning the largest local departures from the smooth spatial field:

```{r sar-map, eval = have_areal, fig.height = 4.2, fig.width = 5.5}
areal.plot(map, LnGM$V$s, title = "V")
```

```{r sar-bf, eval = have_areal}
bayes.factor(LnGM, n.samples = 30, seed = 1)
```

That is the whole custom interface. Give `ngvb_custom()` a function `D(theta)` and a vector
`h`, optionally a prior and a normalizing constant, and any Gaussian model whose precision
factors as $\mathbf D^\top\mathrm{diag}(1/\mathbf h)\mathbf D$ becomes a latent non-Gaussian
one. This is also how you would reach the models listed as "added by hand" earlier, including
the Ornstein-Uhlenbeck process and, through `generic0`, any component whose precision is a
fixed structure matrix times a scale.
