---
title: "Getting started with ngvb2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ngvb2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      eval = requireNamespace("INLA", quietly = TRUE))
```

`ngvb2` extends a latent **Gaussian** model fitted with R-INLA to a latent
**non-Gaussian** model. The Gaussian field has precision $\mathbf{Q}=\mathbf{D}^\top\mathbf{D}$
for a model-specific dependency matrix $\mathbf{D}$; the non-Gaussian version conditions on
mixing variables $\mathbf{V}$,
$$\mathbf{Q}(\boldsymbol\theta,\mathbf{V})=\mathbf{D}(\boldsymbol\theta)^\top\,\mathrm{diag}(1/\mathbf{V})\,\mathbf{D}(\boldsymbol\theta),$$
and fits it by a variational-Bayes loop around INLA. Two functions do the work:
`ng.check(fit)` (is the Gaussian assumption adequate?) and `ngvb(fit)` (fit the extension).

```{r setup}
library(ngvb2)
```

## A random-walk example

We simulate a series with two abrupt level shifts — exactly the kind of local
non-Gaussian behaviour a Gaussian RW1 cannot capture.

```{r sim}
set.seed(3); N <- 100
x <- cumsum(rnorm(N, sd = 0.3)); x[50:N] <- x[50:N] + 6; x[80:N] <- x[80:N] - 9
y <- x + rnorm(N, sd = 0.4)
```

Fit the ordinary latent Gaussian model (remember `config = TRUE`):

```{r lgm}
LGM <- inla(y ~ -1 + f(i, model = "rw1", constr = TRUE),
            data = data.frame(y = y, i = 1:N),
            control.compute = list(config = TRUE))
```

**Check** the latent Gaussian assumption. A small p-value and large per-index
sensitivities `d` flag non-Gaussianity (here at the two shifts):

```{r check}
chk <- ng.check(LGM, compute.fixed = FALSE)
chk
which(chk$components$i$d > quantile(chk$components$i$d, 0.95))
```

**Extend** to the non-Gaussian model — one line, model auto-detected:

```{r fit}
LnGM <- ngvb(LGM, iter = 5, verbose = FALSE)
LnGM
```

The mixing variables `V` are inflated exactly where the series jumps, adding local
flexibility there:

```{r plot, fig.width = 7, fig.height = 3}
plot(LnGM)
```

## Other models

Auto-detection covers `iid`, `rw1`, `rw2`, `ar1`, and `besag`/`besagproper` (CAR/ICAR)
directly from the fit. SAR and SPDE are supplied through `components`:

```{r other, eval = FALSE}
# SAR (W = row-standardised adjacency):
ngvb(LGM, components = list(s = ngvb_operator("sar", W = W)))

# Matern/SPDE (pass the mesh, which the fit does not retain):
ngvb(LGM, components = list(s = ngvb_operator("spde", spde = spde)))
```

## Multiple components

Additive models are handled jointly, each `f()` term getting its own mixing
variables and non-Gaussianity parameter:

```{r multi, eval = FALSE}
LGM  <- inla(y ~ f(time, model = "ar1") + f(subj, model = "iid"),
             data = d, control.compute = list(config = TRUE))
summary(ngvb(LGM))
```
