---
title: "Latent non-Gaussian models with ngvb2"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Latent non-Gaussian models with ngvb2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 3.4, fig.align = "center",
                      eval = requireNamespace("INLA", quietly = TRUE))
have_areal   <- all(vapply(c("spdep", "sf", "spData"), requireNamespace,
                           logical(1), quietly = TRUE)) &&
                requireNamespace("INLA", quietly = TRUE)
have_fmesher <- requireNamespace("fmesher", quietly = TRUE) &&
                requireNamespace("INLA", quietly = TRUE)
set.seed(1)
```

## Overview

A **latent Gaussian model** (LGM) -- the workhorse of R-INLA -- puts a Gaussian prior
on a latent field: a random walk, an autoregression, a spatial effect, a set of random
effects. The Gaussian prior is convenient but rigid: it assumes the *same* variability
everywhere. Real data often breaks that assumption in a few places -- a sudden jump in a
time series, one outlying subject, a sharp boundary between regions, a local hotspot in
space. A Gaussian model must then compromise: either it over-smooths the interesting
feature, or it inflates the variance *everywhere* to fit it.

`ngvb2` relaxes that single assumption. It keeps everything you already have -- the same
INLA model, the same code -- and replaces the Gaussian driving noise by a heavier-tailed
one. The result is a model that **adds flexibility only where the data asks for it** and
falls back to the Gaussian model everywhere else, automatically. Two functions do the work,
both taking an ordinary `inla` fit (run with `control.compute = list(config = TRUE)`):

* **`ng.check(fit)`** -- *should* you relax the Gaussian assumption, and *where*?
* **`ngvb(fit)`** -- fit the non-Gaussian extension.

Both **auto-detect** the model(s) in your fit. Out of the box that covers `iid` random
effects, random walks `rw1` / `rw2`, autoregressions `ar1`, intrinsic areal models
`besag`, and Matern / `SPDE` fields. Anything else -- your own precision matrix -- you
plug in with `ngvb_custom()` (last section).

```{r lib, message = FALSE}
library(ngvb2)
```

## The model in one paragraph

Every LGM writes its latent field $\mathbf x$ through a model-specific **dependency
matrix** $\mathbf D(\boldsymbol\theta)$ that turns $\mathbf x$ into independent Gaussian
increments,
$$\mathbf D(\boldsymbol\theta)\,\mathbf x \;\overset{d}{=}\; \boldsymbol\Lambda,
\qquad \Lambda_i \sim N(0,\,h_i).$$
For a random walk $\Lambda_i=x_{i}-x_{i-1}$; for an autoregression
$\Lambda_i=x_i-\rho x_{i-1}$; and so on. The **non-Gaussian** model keeps the same
$\mathbf D$ but lets each increment have its own variance $V_i$,
$$\Lambda_i \mid V_i \sim N(0,\,V_i),\qquad
  V_i \sim \mathrm{IG}(h_i,\,\eta^{-1}h_i^2),\qquad E[V_i]=h_i .$$
The **mixing variables** $\mathbf V$ are the local flexibility: where $V_i=h_i$ the model
is Gaussian, and where $V_i$ is inflated the increment is allowed to be unusually large.
A single **non-Gaussianity parameter** $\eta\ge 0$ controls how far $\mathbf V$ may stray
from $\mathbf h$; $\eta=0$ is exactly the Gaussian model. Conditional on $\mathbf V$ the
field is still Gaussian, with precision
$$\mathbf Q(\boldsymbol\theta,\mathbf V)=\mathbf D(\boldsymbol\theta)^\top\,
   \mathrm{diag}(1/\mathbf V)\,\mathbf D(\boldsymbol\theta),$$
so `ngvb2` fits the model by a short variational-Bayes loop that repeatedly calls INLA with
$\mathbf V$ fixed and updates $\mathbf V$ and $\eta$. The prior on $\eta$ is an exponential
(a penalised-complexity prior, `alpha.eta`, default rate $1$) that **shrinks toward the
Gaussian model** -- non-Gaussianity has to earn its place. Increase `alpha.eta` for
stronger shrinkage.

## A time series with jumps

`jumpts` is a series that is smooth apart from two abrupt jumps. A Gaussian RW1 assumes
every increment $x_i-x_{i-1}$ has the same variance, so it cannot tell a jump from noise.

```{r rw1-data, fig.height = 3}
plot(jumpts, pch = 19, col = "grey40", main = "jumpts")
```

Fit the ordinary Gaussian RW1, then **check** it:

```{r rw1-check, fig.height = 3.4}
LGM <- inla(y ~ -1 + f(x, model = "rw1"), data = jumpts,
            control.compute = list(config = TRUE))

ng.check(LGM, compute.fixed = FALSE)
```

The left panel is the per-index sensitivity $d_i(y)$: how much the evidence would improve
if increment $i$ were allowed to be non-Gaussian. It **spikes exactly at the two jumps**.
The right panel compares the overall sensitivity $s_0=\sum_i d_i$ (red line) with the
distribution it would have under a truly Gaussian latent field (grey curve): the observed
value sits far in the tail, $p\approx0$. The Gaussian RW1 is inadequate, and we know *where*.

Now **extend** it -- one line, model auto-detected:

```{r rw1-fit, fig.height = 3.2}
LnGM <- ngvb(LGM, iter = 10, verbose = FALSE)      # verbose = TRUE (default) shows a progress bar

plot(LnGM$V$x, type = "h", lwd = 2, col = "grey40",
     xlab = "index", ylab = expression(V[i]), main = "Mixing variables")
```

The mixing variables are inflated precisely at the two jumps and sit near $1$ elsewhere:
the model bought local flexibility exactly where it was needed and stayed Gaussian in
between. The non-Gaussianity parameter is clearly positive:

```{r rw1-summary}
LnGM$eta
```

## Several components: random intercepts and slopes

`Orthodont` records a growth measurement for 27 children at four ages. We give each child
its own **intercept** (baseline) and **slope** (growth rate) -- two `iid` random effects:

```{r ortho-fit}
formula <- value ~ 1 + Female + time + tF +
  f(subject,  model = "iid") +      # random intercept, one per child
  f(subject2, time, model = "iid")  # random slope,    one per child

LGM <- inla(formula, data = Orthodont, control.compute = list(config = TRUE))
```

Why make random effects non-Gaussian? A Gaussian random effect treats the children as
exchangeable draws from a single normal distribution. If one or two children are genuine
outliers -- an unusually high baseline, say -- the Gaussian either shrinks them toward the
mean (hiding the outlier) or widens the distribution for *everyone* to accommodate them.
A non-Gaussian random effect instead keeps a tight distribution for the typical children
and lets the few outliers stand out. `ng.check` reports one diagnostic per component:

```{r ortho-check, fig.height = 6}
ng.check(LGM, compute.fixed = FALSE)
```

```{r ortho-ngvb}
LnGM <- ngvb(LGM, iter = 10, verbose = FALSE)
LnGM$eta          # one non-Gaussianity parameter per component
```

The intercept carries more non-Gaussianity than the slope: a couple of children have
outlying baselines, while the growth rates are close to Gaussian. Nothing was specified by
hand -- the exponential prior on $\eta$ shrank the slope component back to (near-)Gaussian
on its own, and left the intercept free to flag its outliers.

## Areal data: an intrinsic CAR (besag)

Residential burglaries in 49 counties of Columbus, Ohio (`spData`). An intrinsic CAR
(`besag`) spatial effect assumes neighbouring counties are similar, with Gaussian
differences across the edges of the adjacency graph; the non-Gaussian version allows a few
**sharp boundaries** between otherwise-smooth regions.

```{r besag, eval = have_areal, message = FALSE, fig.height = 3.4}
library(sf); library(spdep)
map  <- st_read(system.file("shapes/columbus.gpkg", package = "spData"), quiet = TRUE)
d    <- st_drop_geometry(map[, c("CRIME", "HOVAL", "INC")])
d$s  <- seq_len(nrow(d))
nb   <- poly2nb(map)
Wadj <- Matrix::sparseMatrix(i = rep(seq_along(nb), lengths(nb)), j = unlist(nb), x = 1)

LGM <- inla(CRIME ~ 1 + HOVAL + INC +
              f(s, model = "besag", graph = Wadj, scale.model = TRUE),
            data = d, control.compute = list(config = TRUE))

ng.check(LGM, compute.fixed = FALSE)   # areal model auto-detected
```

```{r besag-fit, eval = have_areal}
LnGM <- ngvb(LGM, iter = 10, verbose = FALSE)
LnGM$eta                                # for besag, V lives on the graph edges
```

## Geostatistical data: a Matern / SPDE field

`weatherdata` (bundled with the package) holds 157 pressure readings across the Pacific
Northwest. We model pressure as a continuous **Matern field** via the SPDE approach --
the standard tool for point-referenced spatial data. A Gaussian Matern smooths the whole
surface uniformly; the non-Gaussian version lets a few locations depart more sharply.
`ngvb2` reads the finite-element operator $\mathbf D=\kappa^2\mathbf C+\mathbf G$
**straight from the fitted `spde` object** -- you pass nothing extra.

```{r spde, eval = have_fmesher, message = FALSE, fig.height = 3.4}
w    <- read.csv(system.file("extdata", "weatherdata.csv", package = "ngvb2"))
mesh <- inla.mesh.2d(loc = w[, c("lon", "lat")], max.edge = c(1.5, 2.5),
                     cutoff = 0.3, max.n.strict = 400)
spde <- inla.spde2.matern(mesh, alpha = 2)
A    <- inla.spde.make.A(mesh, loc = as.matrix(w[, c("lon", "lat")]))
stk  <- inla.stack(tag = "est", data = list(y = w$press), A = list(A),
                   effects = list(s = 1:spde$n.spde))

LGM <- inla(y ~ -1 + f(s, model = spde), data = inla.stack.data(stk),
            control.predictor = list(A = inla.stack.A(stk)),
            control.compute = list(config = TRUE))

ng.check(LGM, compute.fixed = FALSE)   # "SPDE2 model" recovered from the fitted object
```

```{r spde-fit, eval = have_fmesher}
LnGM <- ngvb(LGM, iter = 10, verbose = FALSE)
LnGM$eta
```

`ngvb2` recognises the component because its `f(...)` model is an object built by
`inla.spde2.matern()`: it evaluates that argument from the fit's formula and reads the FEM
matrices `M0`, `M1`, `M2` back from it.

## Your own precision matrix (e.g. a SAR)

If the model is not built in, you supply the dependency matrix
$\mathbf D(\boldsymbol\theta)$ and the constant vector $\mathbf h$ with `ngvb_custom()`.
The engine then builds $\mathbf Q(\mathbf V)=\mathbf D^\top\mathrm{diag}(1/\mathbf V)\mathbf D$,
owns its normalizing constant, and `ng.check()` / `ngvb()` work as usual.

A **simultaneous autoregression** (SAR) for the same Columbus counties uses
$\mathbf D_{\mathrm{SAR}}=\mathbf I-\rho\mathbf W$, with $\mathbf W$ a row-standardised
adjacency and $|\rho|<1$. Unlike the intrinsic besag effect, its mixing variables live on
the counties themselves, so we can **map them**.

```{r sar, eval = have_areal, message = FALSE}
W      <- Matrix::Diagonal(x = 1 / rowSums(Wadj)) %*% Wadj    # row-standardise
N      <- nrow(W)
eigenv <- Re(eigen(as.matrix(W), only.values = TRUE)$values)

# theta = (log precision, logit rho). These functions run in a separate R process,
# so use base R / namespace-qualified calls (exp(t)/(1+exp(t)), not plogis).
op_sar <- ngvb_custom(
  D = function(theta) {
    rho <- exp(theta[2]) / (1 + exp(theta[2]))
    sqrt(exp(theta[1])) * (Matrix::Diagonal(N) - rho * W)
  },
  h = rep(1, N), ntheta = 2,
  logprior = function(theta) {
    rho <- exp(theta[2]) / (1 + exp(theta[2]))
    stats::dnorm(theta[1], 0, 3, log = TRUE) + log(rho) + log(1 - rho)
  },
  lognc = function(theta, Vinv) {                # analytic (stable as rho -> 1)
    rho <- exp(theta[2]) / (1 + exp(theta[2]))
    0.5 * N * theta[1] + sum(log(1 - rho * eigenv)) +
      0.5 * sum(log(Vinv)) - 0.5 * N * log(2 * pi)
  })

# fit the Gaussian SAR through the same engine (V = h), then extend it
LGM  <- inla(CRIME ~ 1 + HOVAL + INC + f(s, model = ngvb_rgeneric(op_sar)),
             data = d, control.compute = list(config = TRUE))
LnGM <- ngvb(LGM, components = list(s = op_sar), iter = 10, verbose = FALSE)
```

```{r sar-map, eval = have_areal, fig.height = 4.2, fig.width = 5}
map$V <- LnGM$V$s
plot(map["V"], pal = function(k) hcl.colors(k, "YlOrRd"), nbreaks = 7,
     main = "SAR mixing weights V")
```

The counties shaded darkest are where the model added the most flexibility -- the local
departures from the smooth spatial field. That is the whole custom interface: give
`ngvb_custom()` a function `D(theta)` and a vector `h`, and any Gaussian model whose
precision factors as $\mathbf D^\top\mathrm{diag}(1/\mathbf h)\mathbf D$ becomes a latent
non-Gaussian one.

## Which models, and adding your own

`ngvb2` applies to any latent Gaussian model whose precision factors as
$$\mathbf Q(\boldsymbol\theta)=\mathbf D(\boldsymbol\theta)^\top\,\mathrm{diag}(1/\mathbf h)\,\mathbf D(\boldsymbol\theta),$$
i.e. models defined by an operator equation $\mathbf D(\boldsymbol\theta)\,\mathbf x = \text{noise}$
that *whitens* the field into independent increments. From a fitted `inla` object it
**recognises automatically** the six most common such models -- `iid`, `rw1`, `rw2`, `ar1`,
`besag` (intrinsic CAR) and `SPDE`/Matern -- which are also among the most-used entries of
`INLA::inla.list.models("latent")`.

**Adding a model that is not built in.** Write down its dependency matrix
$\mathbf D(\boldsymbol\theta)$ -- the operator that turns the field into independent
increments -- and the constant vector $\mathbf h$, and pass them to `ngvb_custom()`. We did
exactly this for the SAR above, where $\mathbf D=\mathbf I-\rho\mathbf W$. Other natural
candidates are AR($p$) ($\mathbf D$ banded with the autoregressive coefficients), seasonal
models, and the Ornstein-Uhlenbeck process. Under the hood `ngvb_custom()` assembles an INLA
`rgeneric` model with precision $\mathbf D^\top\mathrm{diag}(1/\mathbf V)\mathbf D$ and supplies
its normalizing constant, so no C code or hand-written rgeneric is needed -- you provide only
$\mathbf D$ and $\mathbf h$.

**What falls outside the framework.** Models that are not a single operator applied to the
field have no such $\mathbf D$ and are not supported: `bym`/`bym2` (a *sum* of an intrinsic-CAR
and an iid component), proper CAR / `besagproper` (a *conditional* specification
$\mathbf Q=\tau(\mathrm{diag}(d_i)-\rho\mathbf A)$), and coregionalisation or mixture models
(`generic1`/`2`/`3`, `copy`, `z`). For areal data with a spatial-dependence parameter, the SAR
above is the natural operator-based choice.

## Summary

* `ng.check(fit)` tells you whether -- and where -- the latent Gaussian assumption is
  inadequate, straight from the LGM fit.
* `ngvb(fit)` fits the non-Gaussian extension, adding local flexibility only where the data
  supports it and shrinking to the Gaussian model otherwise.
* Six models are recognised automatically; any other operator $\mathbf D$ is supplied through
  `ngvb_custom()`, which wraps INLA's `rgeneric` mechanism.
```
