---
title: "Latent non-Gaussian models with ngvb2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Latent non-Gaussian models with ngvb2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 4,
                      eval = requireNamespace("INLA", quietly = TRUE))
have_areal   <- all(vapply(c("spdep", "sf", "spData"), requireNamespace,
                           logical(1), quietly = TRUE)) &&
                requireNamespace("INLA", quietly = TRUE)
have_fmesher <- requireNamespace("fmesher", quietly = TRUE) &&
                requireNamespace("INLA", quietly = TRUE)
set.seed(1)
```

## The idea

A latent **Gaussian** model (LGM) assumes a latent field $\mathbf{x}$ whose precision
factors through a model-specific dependency matrix $\mathbf{D}(\boldsymbol\theta)$,
$$\mathbf{D}(\boldsymbol\theta)\,\mathbf{x} \;\overset{d}{=}\; \mathbf{Z},
\qquad \mathbf{Z}\sim N(\mathbf 0, \mathrm{diag}(\mathbf h)),$$
so that $\mathbf{Q}(\boldsymbol\theta)=\mathbf{D}^\top\mathrm{diag}(1/\mathbf h)\mathbf{D}$.
The **non-Gaussian** extension replaces the driving noise $\mathbf Z$ by heavier-tailed
noise, which is equivalent to conditioning on mixing variables $\mathbf V$ (with
$E[\mathbf V]=\mathbf h$):
$$\mathbf{Q}(\boldsymbol\theta,\mathbf V)=\mathbf{D}(\boldsymbol\theta)^\top\,
   \mathrm{diag}(1/\mathbf V)\,\mathbf{D}(\boldsymbol\theta).$$
`ngvb2` fits this by a variational-Bayes loop around R-INLA. Two functions do the work,
both taking an ordinary `inla` fit (run with `control.compute = list(config = TRUE)`):

* **`ng.check(fit)`** -- checks whether the latent Gaussian assumption is adequate;
* **`ngvb(fit)`** -- fits the non-Gaussian extension.

Both **auto-detect** the model(s) in the fit (`iid`, `rw1`, `rw2`, `ar1`, `besag`,
and `SPDE`). Anything else you define yourself with `ngvb_custom()` (last section).

```{r lib, message = FALSE}
library(ngvb2)
```

## RW1: a time series with jumps

`jumpts` is a series that is smooth apart from two sudden jumps -- exactly the kind of
feature a Gaussian random walk over-smooths.

```{r rw1}
plot(jumpts, main = "jumpts")

LGM <- inla(y ~ -1 + f(x, model = "rw1"), data = jumpts,
            control.compute = list(config = TRUE))

ng.check(LGM)          # small p-value: the Gaussian RW1 is inadequate
```

```{r rw1-fit}
LnGM <- ngvb(LGM, iter = 8, verbose = FALSE)
plot(LnGM$V$x, type = "h", ylab = "mixing variable V",
     main = "RW1: V flags the two jumps")
```

The mixing variables `V` spike exactly at the jumps, and the non-Gaussianity parameter
`eta` is clearly positive:

```{r rw1-summary}
summary(LnGM)
```

## Several components: random intercepts and slopes

The `Orthodont` growth data uses a random intercept and a random slope, both `iid`.
`ngvb2` detects and extends **both** components jointly, each with its own mixing
variables and non-Gaussianity parameter.

```{r ortho}
formula <- value ~ 1 + Female + time + tF +
  f(subject,  model = "iid") +
  f(subject2, time, model = "iid")

LGM <- inla(formula, data = Orthodont, control.compute = list(config = TRUE))

ng.check(LGM)
LnGM <- ngvb(LGM, iter = 5, verbose = FALSE)
LnGM$eta               # one non-Gaussianity parameter per component
```

## Areal data (ICAR / besag)

Residential burglaries in 49 counties of Columbus, Ohio. We use an intrinsic CAR
(`besag`) spatial effect, whose operator is the graph incidence matrix.

```{r besag, eval = have_areal, message = FALSE}
library(sf); library(spdep)
map  <- st_read(system.file("shapes/columbus.gpkg", package = "spData"), quiet = TRUE)
d    <- st_drop_geometry(map[, c("CRIME", "HOVAL", "INC")])
d$s  <- seq_len(nrow(d))
nb   <- poly2nb(map)
Wadj <- Matrix::sparseMatrix(i = rep(seq_along(nb), lengths(nb)),
                             j = unlist(nb), x = 1)

LGM <- inla(CRIME ~ 1 + HOVAL + INC + f(s, model = "besag", graph = Wadj, scale.model = TRUE),
            data = d, control.compute = list(config = TRUE))

ng.check(LGM)                              # areal model auto-detected
LnGM <- ngvb(LGM, iter = 5, verbose = FALSE)
cat("eta =", round(LnGM$eta, 3), "\n")     # for ICAR, V is one weight per graph edge
```

## Geostatistical data (SPDE / Matern)

For continuous-space Matern fields fitted through the SPDE approach, `ngvb2` recovers the
finite-element operator $\mathbf D = \kappa^2\mathbf C+\mathbf G$ **directly from the fitted
model** -- you pass nothing extra.

```{r spde, eval = have_fmesher, message = FALSE}
n    <- 100
loc  <- matrix(runif(n * 2), n, 2)
mesh <- fmesher::fm_mesh_2d(loc, max.edge = c(0.2, 0.5), cutoff = 0.05)
spde <- inla.spde2.matern(mesh)
A    <- fmesher::fm_basis(mesh, loc)

field <- rnorm(mesh$n, sd = 1.2)
field[sample(mesh$n, 3)] <- field[sample(mesh$n, 3)] + 8    # a few heavy latent values
y     <- as.numeric(A %*% field) + rnorm(n, sd = 0.2)
stk  <- inla.stack(data = list(y = y), A = list(A),
                   effects = list(s = 1:mesh$n), tag = "est")

LGM <- inla(y ~ -1 + f(s, model = spde), data = inla.stack.data(stk),
            control.predictor = list(A = inla.stack.A(stk)),
            control.compute = list(config = TRUE))

ng.check(LGM)                              # "SPDE2 model" recovered from the fit
LnGM <- ngvb(LGM, iter = 5, verbose = FALSE)
cat("eta =", round(LnGM$eta, 3), "\n")
```

`ngvb2` recognises the component because its `f(...)` model is an object produced by
`inla.spde2.matern()`; it reads the FEM matrices `M0`, `M1`, `M2` back from that object.

## Custom precision matrices (e.g. SAR)

For a model that is **not** built in, supply the dependency matrix $\mathbf D(\boldsymbol\theta)$
and the constant vector $\mathbf h$ with `ngvb_custom()`. The engine then builds
$\mathbf Q(\mathbf V)=\mathbf D^\top\mathrm{diag}(1/\mathbf V)\mathbf D$, owns its normalizing
constant, and both `ng.check()` and `ngvb()` work as usual.

A simultaneous autoregression (SAR) has $\mathbf D_{\mathrm{SAR}}=\mathbf I-\rho\mathbf W$, with
$\mathbf W$ a row-standardised adjacency and $|\rho|<1$. (This is also the natural way to put
a *proper* areal effect in this framework: a proper CAR is a conditional specification without
an operator $\mathbf D$, whereas the SAR is defined directly through one.)

```{r sar, eval = have_areal, message = FALSE}
W      <- Matrix::Diagonal(x = 1 / rowSums(Wadj)) %*% Wadj   # row-standardise
N      <- nrow(W)
eigenv <- Re(eigen(as.matrix(W), only.values = TRUE)$values)

# theta = (log precision, logit rho). NB: these functions run in a separate R
# process, so use base R / namespace-qualified calls (no bare plogis / Diagonal).
op_sar <- ngvb_custom(
  D = function(theta) {
    rho <- exp(theta[2]) / (1 + exp(theta[2]))
    sqrt(exp(theta[1])) * (Matrix::Diagonal(N) - rho * W)
  },
  h = rep(1, N), ntheta = 2,
  logprior = function(theta) {
    rho <- exp(theta[2]) / (1 + exp(theta[2]))
    stats::dnorm(theta[1], 0, 3, log = TRUE) + log(rho) + log(1 - rho)
  },
  lognc = function(theta, Vinv) {                # analytic (stable as rho -> 1)
    rho <- exp(theta[2]) / (1 + exp(theta[2]))
    0.5 * N * theta[1] + sum(log(1 - rho * eigenv)) +
      0.5 * sum(log(Vinv)) - 0.5 * N * log(2 * pi)
  })

# fit the Gaussian SAR through the same engine (V = h), then extend
LGM  <- inla(CRIME ~ 1 + HOVAL + INC + f(s, model = ngvb_rgeneric(op_sar)),
             data = d, control.compute = list(config = TRUE))

ng.check(LGM, components = list(s = op_sar))
LnGM <- ngvb(LGM, components = list(s = op_sar), iter = 5, verbose = FALSE)
cat("eta =", round(LnGM$eta, 3), "\n")
```

That is the whole custom interface: give `ngvb_custom()` a function `D(theta)` and a vector
`h`, and any Gaussian model whose precision factors as
$\mathbf D^\top\mathrm{diag}(1/\mathbf h)\mathbf D$ becomes a latent non-Gaussian one.
