---
title: "Random intercepts and slopes"
description: >
  Subject-level random effects in a longitudinal growth study, and what changes when the
  random effects are allowed to be non-Gaussian.
output: rmarkdown::html_document
---

```{r setup, include = FALSE}
source("_common.R")
```

```{r lib}
library(ngvb)
library(ggplot2)
library(INLA)
```

`Orthodont` records a growth measurement for 27 children at four ages, together with a sex
indicator and time coded two ways. A quick look at the raw data frame:

```{r ortho-data, fig.alt = "Growth trajectories for 27 children measured at four ages, one small panel per child.", fig.height = 4}
Orthodont$age <- c(8, 10, 12, 14)[Orthodont$time + 1]

ggplot(Orthodont, aes(x = age, y = value)) +
  geom_line(color = "#007fff") +
  geom_point(shape = 1, color = "#007fff") +
  scale_x_continuous(breaks = c(9, 11, 13)) +
  facet_wrap(~ ifelse(Female == 1, paste0("F", subject), paste0("M", subject)), ncol = 11) +
  theme_bw() +
  theme(panel.spacing = unit(0, "lines"),
        strip.background = element_rect(fill = "#ffe5cc"))
```

Two kinds of child-to-child variation matter here. Each child starts at a different height,
which is variation in the intercept, and each child grows at a slightly different rate, which
is variation in the slope. We give every child its own intercept and its own slope, as two
`iid` random effects:
$$\text{value}_{ij} = \beta_0 + \beta\,\text{covariates}
   + \underbrace{a_{\text{subject}(ij)}}_{\text{random intercept}}
   + \underbrace{b_{\text{subject}(ij)}\,\text{time}_{ij}}_{\text{random slope}}
   + \varepsilon_{ij},$$
with $a_k \sim N(0, 1/\tau_a)$ and $b_k \sim N(0, 1/\tau_b)$ across children.

```{r ortho-inla}
formula <- value ~ 1 + Female + time + tF +
  f(subject,  model = "iid") +      # random intercept, one per child
  f(subject2, time, model = "iid")  # random slope, one per child

LGM <- inla(formula, data = Orthodont, control.compute = list(config = TRUE))
```

Why make the random effects non-Gaussian? A Gaussian random effect treats the children as
exchangeable draws from one normal distribution. If a child or two are genuine outliers,
with an unusually high baseline for example, the Gaussian either shrinks them toward the
mean and hides the outlier, or it widens the distribution for everyone to accommodate them.
A non-Gaussian random effect keeps a tight distribution for the typical children and lets
the few outliers stand out. `ng.check()` reports one diagnostic per component:

```{r ortho-check, fig.alt = "Diagnostics for the random intercept and random slope, both sitting well inside their Gaussian reference distributions.", fig.height = 6}
ng.check(LGM, compute.fixed = FALSE)
```

We see that no substantial evidence of non-Gaussianity was found (large $p$) for both random effects.

```{r ortho-ngvb, fig.alt = "ngvb fit summary for the random intercept and slope: mixing weights, non-Gaussianity posteriors and convergence for each component.", fig.height = 8.5}
LnGM <- ngvb(LGM)
plot(LnGM)
```

After fitting the latent non-Gaussian model we see that the intercept carries more non-Gaussianity than the slope. A couple of children have
outlying baselines, while the growth rates are close to Gaussian.

```{r ortho-bf}
bayes.factor(LnGM, n.samples = 30, seed = 1)
```
