---
title: "Get started with ngvb"
description: >
  What a latent non-Gaussian model is, which INLA components are supported, and a
  first end-to-end example on a time series with two jumps.
output: rmarkdown::html_document
---

```{r setup, include = FALSE}
source("_common.R")
```

```{r lib}
library(ngvb)
library(ggplot2)
library(INLA)     # ngvb uses INLA as a backend; attach it to call inla() directly
```

# Overview

A latent Gaussian model (LGM), the workhorse of R-INLA, places a Gaussian prior on a
latent field: a random walk, an autoregression, a spatial effect, or a set of random
effects. That prior is convenient but rigid, because it assumes Gaussianity
everywhere. Real data often break this assumption in just a few places, such as a sudden
jump in a time series, one outlying subject, a sharp boundary between regions, or a local
hotspot in space. A Gaussian model then has to compromise. It either over-smooths the
interesting feature or inflates the variance everywhere to accommodate it.

`ngvb` relaxes that single assumption. It keeps everything you already have, the same
INLA model and the same code, and replaces the Gaussian driving noise with a heavier-tailed
one. The result adds flexibility only where the data ask for it and falls back to the
Gaussian model everywhere else. Two functions do the work, and both take an ordinary `inla`
fit that was run with `control.compute = list(config = TRUE)`:

* `ng.check(fit)` asks whether you should relax the Gaussian assumption, and where.
* `ngvb(fit)` fits the non-Gaussian extension.

# The model in one paragraph

Every LGM writes its latent field $\mathbf x$ through a model-specific dependency matrix
$\mathbf D(\boldsymbol\theta)$ that turns $\mathbf x$ into independent Gaussian increments,
$$\mathbf D(\boldsymbol\theta)\,\mathbf x \;\overset{d}{=}\; \boldsymbol\Lambda,
\qquad \Lambda_i \sim N(0,\,h_i).$$
For a random walk $\Lambda_i=x_{i}-x_{i-1}$, for an autoregression
$\Lambda_i=x_i-\rho x_{i-1}$, and so on. Equivalently, the Gaussian field has precision
$\mathbf Q(\boldsymbol\theta)=\mathbf D(\boldsymbol\theta)^\top\mathrm{diag}(1/\mathbf h)\,
\mathbf D(\boldsymbol\theta)$. The non-Gaussian model keeps the same
$\mathbf D$ but lets each increment have its own variance $V_i$,
$$\Lambda_i \mid V_i \sim N(0,\,V_i),\qquad
  V_i \sim \mathrm{IG}(h_i,\,\eta^{-1}h_i^2),\qquad E[V_i]=h_i .$$
The mixing variables $\mathbf V$ carry the local flexibility. Where $V_i=h_i$ the model is
Gaussian, and where $V_i$ is inflated the increment is allowed to be unusually large. A
single non-Gaussianity parameter $\eta\ge 0$ controls how far $\mathbf V$ may stray from
$\mathbf h$, and $\eta=0$ is exactly the Gaussian model. Conditional on $\mathbf V$ the
field is still Gaussian, with precision
$$\mathbf Q(\boldsymbol\theta,\mathbf V)=\mathbf D(\boldsymbol\theta)^\top\,
   \mathrm{diag}(1/\mathbf V)\,\mathbf D(\boldsymbol\theta).$$
`ngvb` fits the model with a short variational-Bayes loop that repeatedly calls INLA with
$\mathbf V$ fixed and then updates $\mathbf V$ and $\eta$. The prior on $\eta$ is an
exponential (a penalised-complexity prior with rate `alpha.eta`, default $2$) that shrinks
toward the Gaussian model, so non-Gaussianity has to earn its place. Increase `alpha.eta`
for stronger shrinkage.

Two variational algorithms are available. The default, `method = "SVI"`, is a structured
mean-field scheme that updates $q(\mathbf V)$ and $q(\eta)$ in turn, and it is the more
reliable choice. The alternative, `method = "SCVI"`, integrates $\mathbf V$ out of the
$\eta$ update analytically and so converges in fewer iterations, but its collapsed marginal
can over-shrink a weak yet genuine effect all the way to the Gaussian model.

# Which models and likelihoods are supported

The non-Gaussianity in `ngvb` lives entirely in the latent field, on its driving noise, and never touches the
observation model. So **every INLA likelihood is available**: Gaussian, Poisson, binomial,
negative binomial, survival families, and the rest all work unchanged. You just fit your LGM
as usual and hand it to `ng.check()` or `ngvb()`.

The latent components split into three groups.

**Detected automatically.** If your fit uses one of these, `ngvb()` and `ng.check()` find it
with no extra arguments: `iid`, `rw1`, `rw2`, `ar1`, `besag` (the intrinsic CAR), `spde`
/ `spde2` (Matern fields), and `seasonal`. Each of these is a single operator applied to
independent noise, so it has the dependency matrix $\mathbf D$ the method needs.

**Added by hand.** Some models are not auto-detected but are still described by
$\mathbf D(\boldsymbol\theta)\,\mathbf x \;\overset{d}{=}\; \boldsymbol\Lambda$, so
you can build them with `ngvb_operator()` which expects the matrix $\mathbf D$, the vector
$\mathbf h$. This covers the simultaneous autoregression (`sar`) or  the Ornstein-Uhlenbeck process (`ou`).
It is also possible to use the matrices and operators of the `ngme2` package to define the LnGM;
both are covered in [Custom models](custom-models.html).

**Not compatible.** Some components are not a single operator applied to one field, and the
method does not apply to them. The main cases are the additive spatial models `bym`, `bym2`,
`besag2` (a latent field that is a sum of parts), the conditionally specified `besagproper`,
the copy and mixture constructions `generic1`, `generic2`, `generic3`, `copy`, `scopy`, `z`,
the measurement-error models `mec`, `meb`, and the nonlinear-covariate effects `sigm`,
`revsigm`, `log1exp`, `logdist`. For an additive model such as `bym`, the practical fix is to
split it into its `besag` and `iid` parts as two separate `f()` terms, which `ngvb()` then
extends jointly.

Priors carry over the same way: `pc.prec`, `loggamma`, and `normal` precision priors on the
original `f()` term are read from the fit and reused automatically; anything else falls back
to a default prior with a warning.

# A time series with jumps

`jumpts` is a series that is smooth apart from two abrupt jumps.

```{r rw1-data, fig.alt = "The jumpts series: a smooth trend broken by two abrupt vertical jumps.", fig.height = 3}
ggplot(jumpts, aes(x, y)) +
  geom_line(colour = "grey70") +
  geom_point(colour = "grey30", size = 1.3) +
  labs(title = "jumpts", x = "time", y = "y") +
  theme_minimal(base_size = 12)
```

A first-order random walk models the series through its increments,
$$x_i - x_{i-1} \sim N(0,\,1/\tau),$$
which is the operator $\mathbf D = \sqrt{\tau}\,\mathbf D_0$ with $\mathbf D_0$ the
first-difference matrix. A Gaussian RW1 forces every increment to share the same variance,
so it cannot tell a jump from ordinary noise. Fit it with INLA:

```{r rw1-inla}
LGM <- inla(y ~ -1 + f(x, model = "rw1"), data = jumpts,
            control.compute = list(config = TRUE))
```

Now check whether the latent Gaussian assumption holds:

```{r rw1-check, fig.alt = "Diagnostic for the RW1 fit: per-index sensitivity spiking at the two jumps, and the observed overall sensitivity far in the tail of its Gaussian reference distribution.", fig.height = 3.4}
check <- ng.check(LGM, compute.fixed = FALSE)
```

The left panel is the per-index sensitivity $d_i(y)$, which measures how much the evidence
would improve if increment $i$ were allowed to be non-Gaussian. It spikes at the two jumps.
The right panel compares the observed overall sensitivity $s_0=\sum_i d_i$ (the vertical
line) against the distribution it would have under a truly Gaussian field (the grey curve).
The observed value sits far out in the tail, with $p\approx0$, so the Gaussian RW1 is
inadequate and we know exactly where.

Now we extend it to non-Gaussianity. The model is auto-detected, so this is a single call:

```{r rw1-fit, fig.alt = "ngvb fit summary: mixing weights inflated at the two jumps and near one elsewhere, the posterior of the non-Gaussianity parameter, and its convergence across iterations.", fig.height = 6}
LnGM <- ngvb(LGM)
plot(LnGM)
```

For each component the plot shows the mixing weights $V_i/h_i$ (the posterior mean with a
90 percent credible interval, and a cross at the plug-in value $1/E[1/V_i]$ that INLA is
actually refit with, which sits a little below the mean because $q(V_i)$ is right-skewed). The final panel tracks
$E[\eta]$ across the iterations. The mixing variables are inflated precisely at the two
jumps and sit near one elsewhere, so the model bought local flexibility exactly where it was
needed and stayed Gaussian in between. The non-Gaussianity parameter is clearly positive.

Beyond the point estimate, you can integrate the mixing variables V out entirely by sampling from their variational posterior and refitting the LGM at each draw. This gives the Bayes factor against the Gaussian model, and, pooled with importance weights, posterior summaries of anything the underlying inla() fit reports:

```{r rw1-bf}
samples <- ngvb_sample(LnGM, n.samples = 30)
bayes.factor(samples)               # LnGM vs. LGM marginal-likelihood ratio
summary(samples)                    # importance-weighted summaries, V integrated out
```

On the usual scale a $\log_{10}$ Bayes factor above $0.5$, $1$, or $2$ is substantial, strong,
or decisive evidence for the non-Gaussian model. Here it is far into the decisive range. The
`ess` field is the effective number of draws behind the estimate; if it is small, raise
`n.samples`.

# Where to next

The same two calls carry over to every other kind of data:

* [Geostatistical data](geostatistical.html) fits a Matern (SPDE) field to pressure
  readings and maps, before `ngvb()` is ever run, where the spatial predictions will move.
* [Areal data](areal.html) applies the intrinsic CAR (`besag`) model to burglary rates
  across Columbus, Ohio.
* [Random intercepts and slopes](longitudinal.html) treats subject-level random effects in a
  longitudinal growth study.
* [Custom models](custom-models.html) builds a simultaneous autoregression by hand with
  `ngvb_custom()`, and borrows ready-made operators from the `ngme2` package.
